---
title: "Early detection of changes in COVID-19 incidence using NHS pathways data"
author: "Thibaut Jombart, Dirk Schumacher, Quentin Leclerc, Emily Nightingale, Sam Abbott, Stéphane Ghozzi, Stefan Flasche, Mark Jit, Andre Charlett, Michael Höhle, John Edmunds, CMMID"
date: "`r format(Sys.time(), '%A %d %B %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: zenburn
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_collapse: no
    toc_depth: 4
    toc_float: yes
    css: !expr here::here('css', 'style.css')
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      dpi = 80,
                      fig.height = 4,
                      fig.width = 8,
                      warning = FALSE,
                      message = FALSE,
                      dev = "png")
```



<!-- ======================================================= -->
<!-- ======================================================= -->
<!-- ======================================================= -->

# Data preparation {.tabset .tabset-fade .tabset-pills}

## Outline


* **Load scripts**: loads libraries and useful scripts used in the analyses; all
`.R` files contained in `scripts` at the root of the factory are automatically
loaded

* **Load data**: imports datasets, and may contain some *ad hoc* changes to the
data such as specific data cleaning (not used in other reports), new variables
used in the analyses, etc.



## Load packages

```{r libraries}

library(reportfactory)
library(here)
library(rio) 
library(tidyverse)
library(incidence)
library(distcrete)
library(epitrix)
library(earlyR)
library(projections)
library(linelist)
library(remotes)
library(janitor)
library(kableExtra)
library(DT)
library(AnomalyDetection)
library(surveillance)
library(rsample)
library(yardstick)

```



## Load scripts

These scripts will load:

* all scripts stored as `.R` files inside `/scripts/`
* all scripts stored as `.R` files inside `/src/`

These scripts also contain routines to access the latest clean encrypted data
(see next section). 

```{r load_scripts}

reportfactory::rfh_load_scripts()

```





## Load clean data

We import the latest NHS pathways data:


```{r load_data}

pathways <- import_pathways() %>%
  as_tibble()
pathways

```



## Completion date

We extract the completion date from the NHS Pathways file timestamp:

```{r database_date}

database_date <- attr(pathways, "timestamp")
database_date

```

The **completion date** of the NHS Pathways data is
**`r format(database_date, format = "%A %d %b %Y")`**.



## Auxiliary functions

These are functions which will be used further in the analyses.

Function to estimate the generalised R-squared as the proportion of deviance
explained by a given model:

```{r Rsq}

## Function to calculate R2 for Poisson model
## not adjusted for model complexity but all models have the same DF here

Rsq <- function(x) {
  1 - (x$deviance / x$null.deviance)
}

```


Function to classify the day of the week into *weekend*, *Monday*, and *the
rest*; this is used in models having a *weekday* effect:

```{r day_of_week}

## Fn to add day of week
day_of_week <- function(date) {
  day_of_week <- weekdays(date)
  out <- dplyr::case_when(
    day_of_week %in% c("Saturday", "Sunday") ~ "weekend",
    day_of_week %in% c("Monday") ~ "monday",
    TRUE ~ "rest_of_week"
  )
  out <- factor(out, levels = c("rest_of_week", "monday", "weekend"))
  out
}

```




## Add variables

We add the following variable:

* `day`: an integer representing the number of days from the earliest data
reported, used for modelling purposes; the first day is 0

* `weekday`: a factor indicating which day of the week dates are: `weekend`,
  `monday`, or `rest_of_week`

```{r add_variables}

pathways <- pathways %>% 
  mutate(nhs_region = str_to_title(gsub("_"," ",nhs_region)),
         nhs_region = gsub(" Of ", " of ", nhs_region),
         nhs_region = gsub(" And ", " and ", nhs_region),
         day = as.integer(date - min(date, na.rm = TRUE)),
         weekday = day_of_week(date))

```



## Filter data

We create the following subsets of the data:

* `pathways_recent`: NHS pathways data over the last 4 weeks; note that we need
  to redifine relative days as the timescale has changed

```{r }

first_date <- database_date - 28
pathways_recent <- pathways %>%
  filter(date >= first_date)

```


<!-- ======================================================== -->
<!-- ======================================================== -->
<!-- ======================================================== -->

# Simulated data {.tabset .tabset-fade .tabset-pills}

## Outline

We simulate some case counts and study the behaviour of different approaches for
detecting changes in underlying rates of case incidence.



## Simulator

We simulate the following datasets, all covering 30 days of data and using
Poisson distributions:

1. **stationary**: use a constant rate 50 cases / day
2. **stationary+increase**: use a constant rate 50 for 25 days, followed by an
   exponential increase (*r* = 0.2) from day 26 onwards
3. **decrease+increase**: use an exponentially decreasing rate (*r* = -0.2) from
   a start at 50 for 25 days, followed by an exponential increase (*r* = 0.2)
   from day 26 onwards

```{r simulator}

## Note: this is a quick and dirty simulator only used for a proof of concept

#' @param n_days number of days to run simulations for
#' @param change_at day at which to change (increase / decrease)
#' @param n_start initial number of cases to start with
#' @param r_increase daily growth rate for increase
#' @param r_decrease daily growth rate for decrease

simulate_data <- function(n_days = 30, change_at = 25, n_start = 300,
                          r_increase = 0.1, r_decrease = -0.075) {
  
  dates <- seq_len(n_days)
  n_days_change <- max(1L, n_days - change_at)
  days_change <- seq_len(n_days_change)
  
  sim_stationary <- rpois(n = n_days, lambda = n_start)

  sim_stationary_increase <- rpois(n = change_at, lambda = n_start)
  lambda_increase <- tail(sim_stationary_increase, 1L) * exp(r_increase * days_change)
  sim_stationary_increase <- c(sim_stationary_increase,
                               rpois(n = n_days_change,
                                     lambda = lambda_increase))

  days_decrease <- seq_len(change_at) - 1
  lambda_decrease <- n_start * exp(r_decrease * days_decrease)
  sim_decrease_increase <- rpois(n = change_at, lambda = lambda_decrease)
  lambda_increase <- tail(sim_decrease_increase, 1L) * exp(r_increase * days_change)
  sim_decrease_increase <- c(sim_decrease_increase,
                             rpois(n = n_days_change,
                                   lambda = lambda_increase))
  
  data.frame(dates = dates,
             stationary = sim_stationary,
             stationary_increase = sim_stationary_increase,
             decrease_increase = sim_decrease_increase)

}


```



## Simulated data

```{r simulated_data}

set.seed(1)
sim <- simulate_data()
sim_long <- sim %>%
  pivot_longer(-1, names_to = "simulation", values_to = "cases")

ggplot(sim_long, aes(x = dates, y = cases, color = simulation)) +
  geom_point() +
  geom_line() +
  scale_color_viridis_d(end = 0.8) +
  theme_bw() +
  large_txt +
  theme(legend.position = "bottom") +
  geom_vline(xintercept = 25, lty = 2) +
  labs(title = "Simulated data")

```








<!-- ======================================================== -->
<!-- ======================================================== -->
<!-- ======================================================== -->

# Automatic Selection of Models and Outliers DEtection for Epidemics (ASMODEE) {.tabset .tabset-fade .tabset-pills}

## Outline

This approach does the following:

1. splits the data into a training set ('older' data points) and an test set
  containing the *k* most recent data points, for which we evaluate if they
  represent trend shifts

2. fit a number of models to the training, 

3. evaluate each using cross-validation and retains the best-predicting model
  
4. derive the x-th quantiles of the predictive distribution of the model

5. classify each point outside these quantiles as 'outliers'


  

## Implementation

This is largely inspired by Dirk Schumacher's
[dirky-hack](https://gist.github.com/dirkschumacher/dce2b83276a180375d24eb89b1334ef4). 

In the absence of seasonality, the models retained include:

* constant Poisson model
* Poisson with log-linear day effect
* NegBin with log-linear day effect


Models:

```{r models}

.validate_data <- function(data, vars = c("count")) {
  if (is.null(data)) stop("`data` is NULL")
  if (!is.data.frame(data)) stop("`data` is not a `data.frame`")
  if (!nrow(data)) stop("`data` has 0 rows")
  for (e in vars) {
    if (!e %in% names(data)) {
      msg <- sprintf("missing column in `data`: %s", e)
      stop(msg)
    }
  }
}


## poisson models
model_poisson_constant <- function(data) {
  .validate_data(data)
  glm(data = data, formula = count ~ 1, family = "poisson")
}

model_poisson_time <- function(data) {
  .validate_data(data, c("count", "day"))
  glm(data = data, formula = count ~ 1 + day, family = "poisson")
}

model_poisson_time_weekday <- function(data) {
  .validate_data(data, c("count", "day", "weekday"))
  glm(data = data, formula = count ~ 1 + day + weekday, family = "poisson")
}


## negative binomial models
model_nb_time <- function(data) {
  .validate_data(data, c("count", "day"))
  MASS::glm.nb(data = data, formula = count ~ 1 + day)
}

model_nb_time_weekday <- function(data) {
  .validate_data(data, c("count", "day", "weekday"))
  MASS::glm.nb(data = data, formula = count ~ 1 + day + weekday)
}


## make a list of all models tested
models <- list(
  list(name = "poisson_constant", model = model_poisson_constant),
  list(name = "poisson_time", model = model_poisson_time),
  list(name = "nb_time", model = model_nb_time)
)

```


Functions to obtain RMSE for Jackniffed samples:

```{r rmse_functions}

## auxiliary function to fit a model on training set and get RMSE on testing
## set
get_rmse <- function(data, id_test, model) {
  data_fit <- data[-id_test, ,drop = FALSE]
  data_test <- data[id_test, ,drop = FALSE]
  fitted_model <- model(data_fit)
  predictions <- predict(fitted_model,
                         newdata = data_test,
                         type = "response")
  sqrt(sum(data_test$count - predictions)^2)
}


## applies the function above for all jacknife samples and returns the median
## of all RMSE; see ?rsample::vfold_cv for more options

jack_rmse <- function(data, model) {
  n <- nrow(data)
  all_rmse <- vapply(seq_len(n),
                     function(i) get_rmse(data, id_test = i, model = model$model),
                     numeric(1))

  median(all_rmse, na.rm = TRUE)
}

```

Functions to select 'best' model:

```{r select_model}

#' @param x a `data.frame` containing data to be used in the models
#' 
#' @param models a `list` of models to be evaluated, each formulated as a
#'   function taking a single `data` argument and returning an `lml` or `glm`
#'   output

select_model <- function(x, models) {
  models_mrse_values <- vapply(models,
                        function(model) jack_rmse(x, model),
                        numeric(1))
  models_names <- vapply(models,
                         function(e) e$name,
                         character(1))
  models_results <- data.frame(models_names,
                               models_mrse_values)

  ## return best model
  out <- list(mrse = models_results,
              model = models[[which.min(models_mrse_values)]])
  out
  
}


```


Function to find outliers:

```{r find_outliers}

## Function to find outlier in supplementary observations
find_outliers <- function(fitted_model, newdata, alpha = 0.05) {
  pred_ci <- ciTools::add_pi(tb = newdata, fit = fitted_model,
                             alpha = alpha, names = c("lower", "upper"))
  dplyr::mutate(pred_ci,
                outlier = count < lower | count > upper,
                classif = case_when(
                  count < lower ~ "decrease",
                  count > upper ~ "increase",
                  TRUE ~ "normal"),
                classif = factor(classif,
                                 levels = c("increase", "normal", "decrease")))
}

```



General wrapper, fixed value of *k*:

```{r detect_epichange_fixed_k}

#' Detect trend and outliers for a fixed value of `k`
#'
#' @param x a `data.frame` with all data needed for the models specified in `models`
#'
#' @param models a `list` of models, each component being a named list
#'   containing: `name` as a single `character` string and `model` as a function
#'   with single argument `data`, which return an instance of `lm` or `glm`
#'
#' @param k the number of most recent data points left out of the model fitting
#'   procedure
#'
#' @param alpha the type 1 error rate to be used for the outlier classification;
#'   this corresponds to the expected proportion of data points following the
#'   trend which will be mistakenly classified as outliers; defaults to `0.05`,
#'   corresponding to 5 percent.
#'
#' @param ... a series of named arguments indicating the variables to be used in
#'   the model e.g. `foo = bar` where `foo` is a variable defined in the
#'   models formulas and `bar` is the name the matching column in the dataset;
#'   `character` and `integer` are acceptable to indicate columns

detect_epichange_fixed_k <- function(x, models, k = 1, alpha = 0.05, ...) {

  x <- dplyr::select(x, ..., everything())

  n <- nrow(x)
  x_train <- head(x, n - k)
  x_test <- tail(x, k)

  models_comparison <- select_model(x_train, models)
  best_model <- models_comparison$model
  fit <- best_model$model(x_train)
  diagnostics <- find_outliers(fitted_model = fit, x, alpha = alpha)

  diagnostics <- mutate(diagnostics,
                        type = dplyr::if_else(1:n() <= (n - k),
                                              "train",
                                              "test"))
  ## probability of having at least this number of outliers (or more)
  n_outliers <- sum(diagnostics$outlier)
  p_value <- pbinom(n_outliers,
                    size = nrow(diagnostics),
                    prob = 0.05,
                    lower.tail = FALSE)

  out <- list(all_models = models,
              models_mrse = models_comparison$mrse,
              k = k,
              best_model = best_model,
              model_fit = fit,
              n_outliers = n_outliers,
              p_value = p_value,
              results = diagnostics)
  class(out) <- c("epichange", class(out))
  out
}

```



General wrapper, determines the value of *k* automatically, as the smallest *k*
giving the highest number of outliers. For instance, if we code `o` outliers and
`n` normal points, and observe series for the last 5 points:



```{r detect_epichange}

#' @param x a `data.frame` containing data to be used in the models
#' 
#' @param models a `list` of models to be evaluated, each formulated as a
#'   function taking a single `data` argument and returning an `lml` or `glm`
#'   output
#'
#' @param max_k an `integer` indicating the maximum number of recent days left
#'   out of the trend estimation process
#'
#' @param alpha the type 1 error rate to be used for the outlier classification;
#'   this corresponds to the expected proportion of data points following the
#'   trend which will be mistakenly classified as outliers; defaults to `0.05`,
#'   corresponding to 5 percent.
#'
#' @param ... a series of named arguments indicating the variables to be used in
#'   the model e.g. `foo = bar` where `foo` is a variable defined in the
#'   models formulas and `bar` is the name the matching column in the dataset;
#'   `character` and `integer` are acceptable to indicate columns
#' 
detect_epichange <- function(x, models, max_k = 7, alpha = 0.05, ...) {

  x <- dplyr::select(x, ..., everything())
  
  k_values <- seq_len(max_k)

  list_results <- lapply(k_values, function(k)
    detect_epichange_fixed_k(x, models,
                             k = k,
                             alpha = alpha))

  ## This function aims to find the 'best' `k`, using a score adding two
  ## components:
  ## 
  ## 1. the least outliers in the first n-k points
  ## 2. the most outliers in the last `k` points
  ## 
  ## In case of ties, we retain the solution with the most outliers in the last
  ## `k` point (2nd score)
  ##
  ## @param res the value resturned by `detect_epichange_fixed_k`

  score_k <- function(res) {
    k <- res$k
    res <- res$results
    score_1 <- sum(!res$outlier & (res$type == "train"), na.rm = TRUE)
    score_2 <- sum(res$outlier & (res$type == "test"), na.rm = TRUE)
    c(k, score_1, score_2, score_1 + score_2)
  }

  ## Score each model, reorder by decrease total score and score_2, and get the
  ## first hit
  scores <- t(vapply(list_results, score_k, integer(4)))
  colnames(scores) <- c("k", "score_1", "score_2", "score")
  scores <- dplyr::arrange(as.data.frame(scores),
                           dplyr::desc(score),
                           dplyr::desc(score_2))
  best_k <- scores$k[1]

  ## Return the 'best' model
  list_results[[best_k]]
  
}

```



Plotting method:

```{r plot.epichange}

plot.epichange <- function(x, point_size = 2,
                           col_normal = "#8B8B8C",
                           col_increase = "#CB3355",
                           col_decrease = "#32AB96",
                           guide = TRUE,
                           ...) {

  stop_train <- max(x$results$day) - x$k
  col_model <- "#BBB67E"
  
  scale_classif <- scale_color_manual(
    "Change in trend:",
    values = c(decrease = col_decrease, increase = col_increase, normal = col_normal),
    labels = c(decrease = "Decrease", increase = "Increase", normal = "Same trend"),
    drop = FALSE)

  custom_guide <- if (guide) guide_legend(override.aes = list(size = c(4, 4, 3))) else FALSE
  
  ggplot(x$results, aes(x = day, y = count)) +
    theme_bw() +
    large_txt +
    geom_vline(xintercept = stop_train, linetype = 2) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.4, fill = col_model) +
    #geom_line(aes(y = pred), color = col_model, alpha = 0.8) +
    geom_point(aes(color = classif), size = point_size + x$results$outlier) +
    geom_line(alpha = 0.3) +
    scale_classif +
    theme(legend.position = "bottom") +
    guides(color = custom_guide)
}

```




## Results: stationary

```{r }

res_stationary <- detect_epichange(sim, models, alpha = 0.05,
                                   day = dates, count = stationary)
res_stationary
plot(res_stationary)

```



## Results: stationary + increase

```{r }

res_stationary_increase <- detect_epichange(sim, models, alpha = 0.05,
                                            day = dates, count = "stationary_increase")
res_stationary_increase
plot(res_stationary_increase)

```


## Results: decrease + increase

```{r }

res_decrease_increase <- detect_epichange(sim, models, alpha = 0.05,
                                          day = dates, count = "decrease_increase")
res_decrease_increase
plot(res_decrease_increase)

```







<!-- ======================================================== -->
<!-- ======================================================== -->
<!-- ======================================================== -->

# Application to NHS pathways data {.tabset .tabset-fade .tabset-pills}

## Outline

We apply ASMODEE to the NHS pathways data, using all data in first instance
(111/999 calls and 111-online).


## Models considered

The following models are considered in ASMODEE:

* Poisson GLM constant rate
* Poisson GLM linear with time
* Poisson GLM linear with time with weekday effect (WE / Monday / other days)
* Negative Binomial GLM log-linear with time
* Negative Binomial GLM log-linear with time with weekday effect (WE / Monday / other days)

```{r nhs_models}

## make a list of all models tested
nhs_models <- list(
  list(name = "poisson_constant", model = model_poisson_constant),
  #list(name = "poisson_time", model = model_poisson_time),
  #list(name = "poisson_time_weekday", model = model_poisson_time_weekday),
  list(name = "nb_time", model = model_nb_time),
  list(name = "nb_time_weekday", model = model_nb_time_weekday)
)

```


## Overall results


```{r res_overall}

counts_overall <- pathways_recent %>%
  group_by(date, day, weekday) %>%
  summarise(n = sum(count))

res_overall <- detect_epichange(counts_overall, nhs_models, alpha = 0.05,
                                count = "n")
res_overall
plot(res_overall, point_size = 3) +
  labs(title = "ASMODEE on NHS pathways data",
       subtitle = "(last 4 weeks, including 111/999 calls and 111 online)")

```



## Results by NHS region  {.tabset .tabset-fade .tabset-pills}

```{r res_nhs_region, fig.height = 8, fig.width = 10}

counts_nhs_region <- pathways_recent %>%
  group_by(nhs_region, date, day, weekday) %>%
  summarise(n = sum(count)) %>%
  split(.$nhs_region)

res_nhs_region <- lapply(counts_nhs_region,
                         detect_epichange,
                         nhs_models,
                         alpha = 0.05,
                         count = "n")

plots_nhs_region <- lapply(seq_along(res_nhs_region),
                           function(i)
                             plot(res_nhs_region[[i]], point_size = 1, guide = FALSE) +
                               labs(subtitle = names(res_nhs_region)[i]))
cowplot::plot_grid(plotlist = plots_nhs_region)

```




## Results by CCG  {.tabset .tabset-fade .tabset-pills}

For computational reasons, we only do the analyses on the first 12 CCG. 

```{r res_ccg, fig.height = 8, fig.width = 10}

counts_ccg <- pathways_recent %>%
  group_by(ccg_name, date, day, weekday) %>%
  summarise(n = sum(count)) %>%
  complete(date, fill = list(count = 0)) %>% 
  split(.$ccg_name)

## only first 12 CCG
res_ccg <- lapply(counts_ccg[1:12],
                  detect_epichange,
                  nhs_models,
                  alpha = 0.05,
                  count = "n")

## isolate p-values and output table
p_values <- sapply(res_ccg, function(e) e$p_value) %>%
  stack() %>%
  select(ccg_name = ind, p_value = values) %>%
  arrange(p_value)

p_values %>%
  mutate(p_value = format(p_values$p_value, scientific = TRUE, digits = 2)) %>%           
  dt_tab()

## plots
plots_ccg <- lapply(seq_along(res_ccg),
                           function(i)
                             plot(res_ccg[[i]], point_size = 1, guide = FALSE) +
                               labs(title = names(res_ccg)[i]) +
                               theme(plot.title = element_text(size = 11)))
cowplot::plot_grid(plotlist = plots_ccg)

```









<!-- =======================================================  -->
<!-- =======================================================  -->
<!-- ======================================================= -->

# System information {.tabset .tabset-fade .tabset-pills}

## Outline

The following information documents the system on which the document was
compiled.


## System 

This provides information on the operating system.

```{r system_info}
Sys.info()
```

## R environment

This provides information on the version of R used:

```{r R_session}
R.version
```


## R packages

This provides information on the packages used:

```{r R_pkg}
sessionInfo()
```
